Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization

### 2020-08-12 MSc_Dissertation_B156476.tex
Added section on MF comparative research, and completed research gaps and questions

\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}

@@

\begin{mdframed}
\begin{center}
\huge
\vspace{0.3in}
\bf
Disentangling patterns of gene expression in high grade serous ovarian cancer
\vspace{0.2in}
\end{center}
\vspace{0.2in}
\end{mdframed}

@@

There is substantial confusion and lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it takes an effort to figure out in which way ICA was applied" \cite{Sompairac2019}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined. The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.  
According to Cantini \etal\, references \cite{Au-Yeung2014}, \cite{Kairov2017}, \cite{Kong2008} and \cite{Lee2003} optimise metagenes, while references \cite{Meng2016} and \cite{Barillot2013} optimises metasamples.

@@

\subsection{Research goals}
This work is concerned with analysing the patterns gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.

The following research questions were set out early in the project. \emph{TODO: tweak these goals!}
\begin{enumerate}
\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
\item Which of PCA, ICA and NMF is best suited to this analysis?
\item For identified genes, what is the underlying biology?
\item Can we discriminate subtypes of HGSOC from expression data?
\item Is there value in incorporating somatic and germline genomic data?
\end{enumerate}

@@

\subsubsection{Method outline}
An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
\item
Investigation of batch effects by repeating of metagene extraction and heatmaps on the \emph{combined} AOCS and TCGA datasets.
\end{enumerate}

@@

\subsection{Metagene extraction by cluster coherence}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

 for factorizer in NMF, ICA, PCA:
    for k in 2..10:
       Ws = []
       for j in 1..50:
          X = bootstrap sample N from N
          W, H = factorizer(X, rank=k, seed=j)
          append W to Ws

@@

\[
   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}.
\]
The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a heatmap.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
\caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
\label{fig-heatmap-BOTH}
\end{center}
\end{figure}

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
\caption{Boxplots of metasamples from the combined AOCS + TCGA datasets, showing relationship between metasamples dataset origin; Which=0 or 1 for AOCS or TCGA respectively.}
\end{center}
\label{fig-wgd-boxplots-BOTH}
\end{figure}

@@

This work has thrown up several possible lines of further study:
\begin{enumerate}
\item
{\bf Algorithm hyper-parameters}.  NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.  For example, NMF has parameters to encourage sparsity through L1 regularization.   This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
\item
{\bf Selecting components from several factorization ranks}.
In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.  Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.  Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.  Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.   In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
\item
{\bf Cross-dataset factorization stability}.  We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.   A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.  In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).  The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
\item
{\bf Contrasting gene expression patterns between cancers}.  The current work set out to find patters of gene expression in HGSOC.  Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.  One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.  That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
\item
{\bf Systematic comparison with published gene expression patterns}.  In analysing the meaning of each metagene above, some tentative links were made with the research literature.   But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!  A more systematic and ideally automated approach is required.  The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.

@@

\subsection{Gene enrichment raw results}
\label{tab-go-enrichment-results}
TODO
\subsection{Software libraries and versions}
TODO
\label{sec-software-versions}

>>>>

\documentclass[draft,tikz, 12pt,a4paper,oneside,fleqn]{article}

@@

\begin{mdframed}
\begin{center}
\huge
\vspace{0.3in}
\bf
Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization
\vspace{0.2in}
\end{center}
\vspace{0.2in}
\end{mdframed}

@@

There is substantial confusion and lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it takes an effort to figure out in which way ICA was applied" \cite{Sompairac2019}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined. The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.  
According to \cite{Cantini2019}, references \cite{Au-Yeung2014,Kairov2017,Kong2008,Lee2003},  optimise metagenes, while references \cite{Meng2016,Barillot2013} optimises metasamples.

@@

\subsubsection{Comparative studies of dimensionality reduction methods for gene expression analysis}

Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.  They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders  and variational autoencoders .  Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.   Their conclusion was that no single method was optimal, but best performance -- judged by capturing the most biological pathway-associated features -- was achieved by taking the extracted latent dimensions of all the methods, an approach they call the ``BioBombe''.

Meng \etal\ \cite{Meng2016} extensively review dimensionality reduction methods applied to multi-omic datasets.   PCA, ICA and NMF are discussed, but the main focus is on less well known algorithms applicable to multiple datasets, including multiple coinertia analysis and consensus PCA.  The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a common and manageable representation space.

The question of reproducibility across datasets of matrix factorization methods is considered by Cantini \etal. \cite{Cantini2019}.  They experiment with several MF methods applied to a total of 26 cancer transcriptomic datasets (including ovarian).  They conclude the best approach to be ICA, stabilized by the use of Reciprocally Best Hit graphs; in essence requiring reciprocal identification of metagenes between at least two datasets.

\subsection{Research gaps and questions}
Considering the foregoing review of the literature, there has clearly been much work by large teams of researchers directed at understanding the gene expression patterns in HGSOC, as one would expect given its clinical importance.  
Dimensionality reduction is a huge and active field -- the bread-and-butter of machine learning research. 
As we have seen, conclusions both on the underlying biology and the optimal methodology are far from consistent, however.   Some approaches -- such as the BioBombe reviewed above -- are complex with results likely to be difficult to reproduce.

Robustness of research results is of course fundamental.  MF methods offer a means of \emph{unsupervised} learning on transcriptomic datasets to identify key genes active in a disease process.  Would two teams working with identical data and using similar MF methods obtain similar metagenes and so arrive at the same biological conclusions?   The \emph{stability} of MF methods is well considered by many authors, although this has typically been focussed on stability of the underlying algorithm -- NMF and ICA are stochastic in nature.   Much less considered is stability in the presence of \emph{sampling error} which inevitably exists particularly in small datasets.

These and wider considerations suggest the following research questions.

\begin{enumerate}
\item Which are the influential genes, or groups of genes, active in HGSOC according to gene expression data?   
\item How do these patterns of gene expression relate to underlying biological processes.  Do these confirm published results?
\item Which MF method is best suited to this analysis and how should they be applied to achieve robust results in the presence of limited dataset size?
\item Do MF methods yield factors which are predictive of patient survival?
\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
\end{enumerate}

@@

\subsubsection{Methods}
An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.

These steps are detailed in the following sections.

@@

\subsection{Metagene extraction by cluster coherence under bootstrap resampling}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

 for factorizer in NMF, ICA, PCA:
    for k in 2..10:
       Ws = []
       for j in 1..50:
          X = bootstrap sample N from N with replacement
          W, H = factorizer(X, rank=k, seed=j)
          append W to Ws

@@

\[
   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}.
\]
The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a square heatmap.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5
, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
\caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
\label{fig-heatmap-BOTH}
\end{center}
\end{figure}

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
\caption{Boxplots of metasamples from the combined AOCS + TCGA datasets, showing relationship between metasamples dataset origin; Which=0 or 1 for AOCS or TCGA respectively.}
\end{center}
\label{fig-wgd-boxplots-BOTH}
\end{figure}

@@

This work has thrown up several possible lines of further study:
\begin{enumerate}
\item
{\bf Algorithm hyper-parameters}.  NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.  For example, NMF has parameters to encourage sparsity through L1 regularization.   This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
\item
{\bf Selecting components from several factorization ranks}.
In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.  Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.  Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.  Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.   In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.  In fact this is similar to the approach of \cite{WayGregory2019} mentioned in section \ref{sec-comparative-dimensionality-studies}.
\item
{\bf Cross-dataset factorization stability}.  We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.   A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.  In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).  The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
\item
{\bf Contrasting gene expression patterns between cancers}.  The current work set out to find patters of gene expression in HGSOC.  Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.  One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.  That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
\item
{\bf Systematic comparison with published gene expression patterns}.  In analysing the meaning of each metagene above, some tentative links were made with the research literature.   But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!  A more systematic and ideally automated approach is required.  The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.

@@

\subsection{Gene enrichment raw results}
\label{tab-go-enrichment-results}
TODO

\label{sec-software-versions}
